Synthesizing Random Generators via Higher-Order Representations

### 2019-01-16 paper/main.tex
WIP

\begin{enumerate}
\item The target code behaves differently over inputs matching very specific
  patterns.
\item The target code yields some of the responsability of preserving its
  invariants to the functions of its abstract interface.
\end{enumerate}

@@

If we use \megadeth to obtain a random generator for \ensuremath{\Conid{Exp}}, we obtain a uniform
generation probability distribution over constructors, i.e., $P(\ensuremath{\Conid{Val}}) =
P(\ensuremath{\Conid{Add}}) = P(\ensuremath{\Conid{Mul}}) = 1/3$.
%
In this setting, $P(match(foo\#1))$ results $1/2430000$, meaning that, in
average, we will need to generate over than two million test cases in order to
be able to test the first clause of \ensuremath{\Varid{foo}} only once.
%
This situation can be somewhat improved if we decide to use \dragen to obtain a
random generator.
%
By optimizing the generation probabilities in order to benefit the generation of
\ensuremath{\Conid{Add}} constructor in a ratio of $20:1$ with respect to the rest of them, we
obtained a distribution of values where $P(match(foo\#1)) \sim 1/300000$.
%
Although this certainly improves the probability of generating a matching value,
this probability is not substantial enough to become practical.
%
Additionally, by favoring the generation probabilities towards the \ensuremath{\Conid{Add}}
constructor, we found that the probability of generating a value matching the
second clause of \ensuremath{\Varid{foo}} (which does not matches against it) also gets diminished
into an impractival value.

@@

This combinator is evaluated simply by piping the result from the inner
representation.
%
It does not change the evaluation semantics, as it is only considered at
generation time:

@@

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{type}\;\text{HRep}_{Exp}{}\<[16]%
\>[16]{}\mathrel{=}\text{HRep}_{Val}\;\!\oplus\!\;\text{HRep}_{Add}\;\!\oplus\!\;\text{HRep}_{Mul}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{data}\;(\Varid{f}\;\otimes\;\Varid{n})\;\Varid{a}\mathrel{=}\Conid{TagFreq}\;(\Varid{f}\;\Varid{a}){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mathbf{instance}\;(\Varid{f}\;\otimes\;\Varid{n}\;\Downarrow\;\Conid{Exp})\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{step}\;(\Conid{TagFreq}\;\Varid{f})\mathrel{=}\Varid{step}\;\Varid{f}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

@@

\begin{figure}[b]
  \centering
  \input{tikz/hrep.tex}
  \caption{The idea}
\end{figure}

>>>>

\begin{enumerate}
\item The target code behaves differently over inputs matching very specific
  patterns.
\item The target code yields part of the responsability of preserving its
  invariants to the functions of its abstract interface.
\end{enumerate}

@@

If we use \megadeth to automatically derive a random generator for \ensuremath{\Conid{Exp}}, we
obtain a uniform generation probability distribution over constructors, i.e.,
$P(\ensuremath{\Conid{Val}}) = P(\ensuremath{\Conid{Add}}) = P(\ensuremath{\Conid{Mul}}) = 1/3$.
%
In this setting, $P(match(foo\#1))$ results $1/2430000$, meaning that, in
average, we will need to generate over than two million test cases in order to
be able to test the first clause of \ensuremath{\Varid{foo}} only once.
%
This situation can be somewhat improved if we decide to use \dragen to obtain a
random generator.
%
By optimizing the generation probabilities in order to benefit the generation of
\ensuremath{\Conid{Add}} constructor in a ratio of $20:1$ with respect to the rest of them, we
obtained a distribution of values where $P(match(foo\#1)) \approx 1/300000$.

%
Although this certainly improves the probability of generating a matching value,
this probability is not substantial enough to become practical.
%
Additionally, by favoring the generation probabilities towards the \ensuremath{\Conid{Add}}
constructor, we found that the probability of generating a value matching the
second clause of \ensuremath{\Varid{foo}} (which does not matches against it) also gets diminished
into an impractival value.

@@

This combinator is evaluated back to our target data type simply by piping the
result from the inner representation.
%
It does not change the evaluation semantics, as it is only considered at
generation time:

@@

With the introduced combinators, we can easily create a type synonym \ensuremath{\text{HRep}_{Exp}}
to refer to the canonical representation of our original data type \ensuremath{\Conid{Exp}},
tagging for instance the representation of the constuctor \ensuremath{\Conid{Add}} to be generated
in double the proportion of rest of the representations:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
\column{24}{@{}>{\hspre}l<{\hspost}@{}}%
\column{34}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{type}\;\text{HRep}_{Exp}{}\<[16]%
\>[16]{}\mathrel{=}{}\<[24]%
\>[24]{}\text{HRep}_{Val}{}\<[E]%
\\
\>[16]{}\,\oplus\;{}\<[24]%
\>[24]{}\text{HRep}_{Add}\;{}\<[34]%
\>[34]{}\otimes\;\mathrm{2}{}\<[E]%
\\
\>[16]{}\,\oplus\;{}\<[24]%
\>[24]{}\text{HRep}_{Mul}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

@@

This previous definition can be interpreted graphically as it is shown in the
Figure \ref{fig:hrep}.
%
Curly arrows represent the structural information extracted using
meta-programming.

\begin{figure}[t]
  \centering
  \input{tikz/hrep.tex}
  \caption{Higher order representation of the data type \ensuremath{\Conid{Exp}}, using structural
    information from the function \ensuremath{\Varid{foo}} and the abstract interface of the module
    \ensuremath{\Conid{M}}.}
  \label{fig:hrep}
\end{figure}

### 2019-01-16 paper/sections/hrep.tex
WIP

\begin{code}
type HRep_Exp  = HRep_Val oplus'' HRep_Add oplus'' HRep_Mul
\end{code}

\begin{code}
data (f otimes n) a = TagFreq (f a)

instance (f otimes n down Exp) where
  step (TagFreq f) = step f
\end{code}

@@

\begin{figure}[b]
  \centering
  \input{tikz/hrep.tex}
  \caption{The idea}
\end{figure}

Pellentesque dapibus suscipit ligula. Donec posuere augue in quam. Etiam vel
tortor sodales tellus ultricies commodo. Suspendisse potenti. Aenean in sem ac
leo mollis blandit. Donec neque quam, dignissim in, mollis nec, sagittis eu,
wisi. Phasellus lacus. Etiam laoreet quam sed arcu. Phasellus at dui in ligula
mollis ultricies. Integer placerat tristique nisl. Praesent augue. Fusce
commodo. Vestibulum convallis, lorem a tempus semper, dui dui euismod elit,
vitae placerat urna tortor vitae lacus. Nullam libero mauris, consequat quis,
varius et, dictum id, arcu. Mauris mollis tincidunt felis. Aliquam feugiat
tellus ut neque. Nulla facilisis, risus a rhoncus fermentum, tellus tellus
lacinia purus, et dictum nunc justo sit amet elit.

Nullam eu ante vel est convallis dignissim. Fusce suscipit, wisi nec facilisis
facilisis, est dui fermentum leo, quis tempor ligula erat quis odio. Nunc porta
vulputate tellus. Nunc rutrum turpis sed pede. Sed bibendum. Aliquam posuere.
Nunc aliquet, augue nec adipiscing interdum, lacus tellus malesuada massa, quis
varius mi purus non odio. Pellentesque condimentum, magna ut suscipit hendrerit,
ipsum augue ornare nulla, non luctus diam neque sit amet urna. Curabitur
vulputate vestibulum lorem. Fusce sagittis, libero non molestie mollis, magna
orci ultrices dolor, at vulputate neque nulla lacinia eros. Sed id ligula quis
est convallis tempor. Curabitur lacinia pulvinar nibh. Nam a sapien.

>>>>

With the introduced combinators, we can easily create a type synonym |HRep_Exp|
to refer to the canonical representation of our original data type |Exp|,
tagging for instance the representation of the constuctor |Add| to be generated
in double the proportion of rest of the representations:

\begin{code}
type HRep_Exp  =       HRep_Val
               oplus'  HRep_Add  otimes 2
               oplus'  HRep_Mul
\end{code}

@@

This previous definition can be interpreted graphically as it is shown in the
Figure \ref{fig:hrep}.
%
Curly arrows represent the structural information extracted using
meta-programming.

\begin{figure}[t]
  \centering
  \input{tikz/hrep.tex}
  \caption{Higher order representation of the data type |Exp|, using structural
    information from the function |foo| and the abstract interface of the module
    |M|.}
  \label{fig:hrep}
\end{figure}

### 2019-01-16 paper/sections/randomtesting.tex
WIP

If we use \megadeth to obtain a random generator for |Exp|, we obtain a uniform
generation probability distribution over constructors, i.e., $P(|Val|) =
P(|Add|) = P(|Mul|) = 1/3$.
%
In this setting, $P(match(foo\#1))$ results $1/2430000$, meaning that, in
average, we will need to generate over than two million test cases in order to
be able to test the first clause of |foo| only once.
%
This situation can be somewhat improved if we decide to use \dragen to obtain a
random generator.
%
By optimizing the generation probabilities in order to benefit the generation of
|Add| constructor in a ratio of $20:1$ with respect to the rest of them, we
obtained a distribution of values where $P(match(foo\#1)) \sim 1/300000$.
%
Although this certainly improves the probability of generating a matching value,
this probability is not substantial enough to become practical.
%
Additionally, by favoring the generation probabilities towards the |Add|
constructor, we found that the probability of generating a value matching the
second clause of |foo| (which does not matches against it) also gets diminished
into an impractival value.

>>>>

If we use \megadeth to automatically derive a random generator for |Exp|, we
obtain a uniform generation probability distribution over constructors, i.e.,
$P(|Val|) = P(|Add|) = P(|Mul|) = 1/3$.
%
In this setting, $P(match(foo\#1))$ results $1/2430000$, meaning that, in
average, we will need to generate over than two million test cases in order to
be able to test the first clause of |foo| only once.
%
This situation can be somewhat improved if we decide to use \dragen to obtain a
random generator.
%
By optimizing the generation probabilities in order to benefit the generation of
|Add| constructor in a ratio of $20:1$ with respect to the rest of them, we
obtained a distribution of values where $P(match(foo\#1)) \approx 1/300000$.

%
Although this certainly improves the probability of generating a matching value,
this probability is not substantial enough to become practical.
%
Additionally, by favoring the generation probabilities towards the |Add|
constructor, we found that the probability of generating a value matching the
second clause of |foo| (which does not matches against it) also gets diminished
into an impractival value.
